Co-evolution of lexical meaning \& pragmatic use

### 2018-03-17 paper/cogsci-submission2.tex
fixes up to sec 3

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}
Types are what evolution operates on. A type is what determines an agent's behavior and thereby
an agent's fitness, so that it makes sense to speak of the fitness of a type itself. To study
the joint evolution of semantic meaning and pragmatic use, types are pairs consisting of a
lexicon and a linguistic strategy of language use.

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a set of
states. A linguistic behavior specifies a probabilistic speaker rule (a probabilistic choice of
message for each state) and a probabilistic hearer rule (a probabilistic choice of state for
each message) given a lexicon. There are many ways of making these general notions
concrete. In the following we aim at a general formulation of speaker and listener
behavior which is as simple as possible but still in line with both a game-theoretic tradition
\citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and a Bayesian
tradition
\citep[e.g.,][]{frank+goodman:2012,goodman+stuhlmueller:2013,GoodmanFrank2016:Pragmatic-Langu}.

@@

Linguistic strategies define dispositions to produce and interpret messages given a lexicon. We
distinguish between two kinds strategies. {\em Literal interlocutors} produce and
interpret messages literally, being guided only by their lexica. {\em Pragmatic interlocutors}
instead engage in mutual reasoning to inform their choices. Recent probabilistic models of
rational language use
\citep{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. The hierarchy's bottom,
level $0$, corresponds to literal language use, as in Equations~(\ref{h:level0}) and
(\ref{s:level0}). Pragmatic language users of level $n + 1$ act (approximately) rational with
respect to level-$n$ behavior of their interlocutors, as in Equations~(\ref{h:leveln}) and
(\ref{s:leveln}).
\begin{alignat}{3}
&H_{0}(s \mid m;L) &&\propto pr(s) L_{[s,m]} \label{h:level0}\\
&S_{0}(m \mid s;L) &&\propto \exp(\lambda \; L_{[s,m]}) \label{s:level0}\\
&H_{n+1}(s|m;L) &&\propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) &&\propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{alignat}

According to (\ref{h:level0}), a literal hearer's interpretation of a message depends on her
lexicon and her prior over states, $pr \in \Delta(S)$, which is here assumed flat for
simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
message according to their lexicon. Pragmatic hearers, defined in (\ref{h:leveln}), instead use
Bayes' rule to weigh interpretations based on a conjecture about speaker behavior. Speaker
behavior is regulated by a soft-max parameter $\lambda \geq 0$
\citep{luce:1959,sutton+barto:1998}. As $\lambda$ increases, choices approximate strict
maximization of expected utilities. Expected utility of a message $\messg$ in state $\state$
for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer
will assign to or choose the correct meaning.\footnote{An alternative choice for the speaker's
  utility function is $\log H_{n}(s|m;L)$. Conceptually, the difference is that without the
  logarithm, utilities are derived from action choices, while with the logarithm utilities are
  derived from how closely the listener's beliefs (after hearing an utterance) align with the
  speaker's, as measure by Kullback-Leibler divergence
  \cite{frank+goodman:2012,qing+franke:2015}. In an evolutionary context, where fitness comes
  from actual interactions with the environment, not just beliefs held about it, the
  formulation without the logarithm may be slightly more appealing. Moreover, log-based speaker
  utilities cause the speaker never to use false utterances. Since we aim here for a situation
  in which any type has a reasonable chance to be mistaken for any other based on their
  production behavior (see Section~\ref{sec:learnability}), we want a definition of speaker
  types which also send any message in any state with positive probability.} For literal
speakers, utility only tracks truthfulness. Literal speakers choose any true message with equal
probability but may send false messages as well with a probability dependent on $\lambda$.

@@

In contrast to their literal counterparts of level $0$, pragmatic agents of level $1$ who use
$\Llack$ associate $\msome$ preferentially with $\ssome$. This association is not perfect, and
usually less strong than what agents with a lexicalized upper bound in $\Lbound$ can achieve --
with or without pragmatic reasoning. Higher order reasoning beyond level $1$ leads to stronger
associations of $\msome$ and $\ssome$ also for the receiver. Still, the case study presented in
Section~\ref{sec:si-case-study} will consider sender and receiver behavior at levels 0 and 1,
as the latter are the simplest pragmatic reasoning types which show a tendency to
communicatively attuned pragmatic enrichment. Using only level 1 reasoning and possibly small
$\lambda$ is therefore a conservative choice that works \emph{against} the fitness-based
selection of pragmatic language use for a notion of fitness defined as communicative success,
which is introduced next.

@@

Languages are shaped not only by functionalist forces towards greater communicative
success. Another important factor is the fidelity with which linguistic knowledge is
transmitted. In biological evolution, where types are expressed genetically, transmission
infidelity comes into the picture through infrequent and mostly random mutation and genetic
drift \citep{kimura:1983}. However, an agent's lexicon and a disposition for pragmatic
reasoning are likely not inherited genetically. They need to be learned from observation, such
as in conditional imitation (see Appendix~\ref{sec:deriv-repl-mutat}).  Concretely, when agents
learn from or imitate type $j$, they observe the overt linguistic behavior of type $j$ and need
to infer the covert type that most likely produced the observed behavior.

@@

Following \citet{griffiths+kalish:2005,griffiths+kalish:2007} we model language acquisition as
a process of Bayesian inference in which learners combine the likelihood of a type producing
the witnessed learning input with prior inductive biases. In a Bayesian setting these biases
can be codified in a prior $P \in \Delta(T)$, which reflects the amount of data a learner
requires to faithfully acquire the language of the teacher
\citep[cf.][450]{griffiths+kalish:2007}. Put differently, types favored by the prior are easier
to learn because they require less data than less favored types to be inferred, even if they
are both compatible with the learning data at hand.

The extent of the prior's influence has been shown to heavily depend on the learning strategy
assumed to underlie the inference process. Early simulation results suggested that weak biases
could be magnified by exposing learners to only small data samples
\citep[e.g. in][]{brighton:2002}. \citet{griffiths+kalish:2007} showed that iterated learning
converges to the prior in the limit, i.e., that the resulting distribution over types
corresponds to the learners' prior distribution and is not influenced by the amount of input
given to them, if learners select random hypothesis from their posterior beliefs about which
types may have generated the learning data that they saw. More deterministic strategies such as
the adoption of the type with the highest posterior probability, so-called {\it maximum a
  posterior estimation} (MAP), increase the influence of both the prior and the data
\citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we use a parameter $l\ge1$ to
modulate between posterior sampling and the MAP strategy. When $l = 1$ learners sample from the
posterior. The learners' propensity to maximize the posterior grows as $l$
increases.\footnote{The way the posterior over types is formed and how learners employ it are
  important design choices. There are other alternatives as well. For instance, instead of
  adopting a single type, learners could keep their uncertainty over types and base their
  production behavior on the posterior over types they form \citep{burkett+griffiths:2010}. A
  straightforward operationalization of this idea would then see production defined as a
  two-step process wherein a speaker first samples from her posterior and then produces
  utterances based on the type she sampled, as done in \citet{kirby+etal:2015}. Beyond adding
  complexity to the model, a disadvantage of this approach is that it could happen that a
  speaker that first uttered a Hungarian utterance speaks Quechua in the next, so to speak. We
  therefore maintain the simpler assumption of a single type being acquired in the present
  context, and postpone further investigation of more complex type-adoption schemes to future
  research.}

@@

When $l=1$ learners sample a type to adopt from their posterior distribution which they obtain
from observing the teacher produce utterances in particular states. From
\citet{griffiths+kalish:2007}, we know that under $l=1$ iterated Bayesian learning will
converge to a population proportion that reflects that prior over types. If $l>1$ learners tend
to pick types which are \emph{a posteriori} more likely with a higher probability than less
likely types. Even in the absence of a prior bias, iterated Bayesian learning with $l>1$ can
lead to populations in which a particular type $t_i$ has a higher proportion than other
types. For $l>1$, much hinges on particular asymmetries in production likelihoods
$P(d \mid t_j)$, and therefore on the composition of the set of types in general. Intuitively
put, if there are many types $t_j$ which all happen to produce data that is most likely to be
produced by $t_i$, this can lead to a larger proportion of $t_i$ in the long run even when
priors over types are uniform. Appendix~\ref{sec:example-iter-bayes} provides an abstract
example.

>>>>

\subsubsection{Types: Lexica and pragmatic strategies}
\label{sec:languages+use}
Types are what evolution promotes or demotes. A type determines an agent's behavior and thereby
an agent's fitness, so that it makes sense to speak of the fitness of a type itself. To study
the joint evolution of semantic meaning and pragmatic use, types are pairs consisting of a
lexicon and a linguistic strategy of language use.

Agents play signaling games, in which the speaker wants to communicate a world state $s$ with a
message $m$ to a hearer who receives $m$ but does not know $s$
\citep[e.g.][]{lewis:1969,Skyrms2010:Signals}. A lexicon associates each message with a set of
states. A linguistic behavior specifies a probabilistic speaker rule (a probabilistic choice of
message for each state) and a probabilistic hearer rule (a probabilistic choice of state for
each message) given a lexicon.

@@

Linguistic strategies define dispositions to produce and interpret messages given a lexicon. We
distinguish between two kinds strategies. {\em Literal interlocutors} produce and interpret
messages literally, being guided only by their lexica. {\em Pragmatic interlocutors} instead
engage in mutual reasoning to inform their choices. Recent game-theoretic
\citep[e.g.,][]{benz:2006,benz+vRooij:2007,jaeger:2007b,franke+jaeger:2014} and probabilistic
models of rational language use
\citep[e.g.,][]{frank+goodman:2012,FrankeJager2015:Probabilistic-p,GoodmanFrank2016:Pragmatic-Langu}
capture different types of pragmatic behavior in a reasoning hierarchy. In the following we aim
at a general formulation of speaker and listener behavior which is as simple and as practical
as possible for our current purposes, but
still in line with both the game-theoretic and the Bayesian traditions.

The hierarchy's bottom, level $0$, corresponds to literal language use, as in
Equations~(\ref{h:level0}) and (\ref{s:level0}). Pragmatic language users of level $n + 1$ act
(approximately) rational with respect to level-$n$ behavior of their interlocutors, as in
Equations~(\ref{h:leveln}) and (\ref{s:leveln}).
\begin{alignat}{3}
&H_{0}(s \mid m;L) &&\propto pr(s) L_{[s,m]} \label{h:level0}\\
&S_{0}(m \mid s;L) &&\propto \exp(\lambda \; L_{[s,m]}) \label{s:level0}\\
&H_{n+1}(s|m;L) &&\propto pr(s) S_{n}(m|s;L) \label{h:leveln}\\
&S_{n+1}(m|s;L) &&\propto  \exp(\lambda \; H_{n}(s|m;L)) \label{s:leveln}
\end{alignat}

According to (\ref{h:level0}), a literal hearer's interpretation of a message depends on her
lexicon and her prior over states, $pr \in \Delta(S)$, which is here assumed flat for
simplicity. Literal interpreters thereby choose an arbitrary true interpretation for each
message according to their lexicon. Pragmatic hearers, defined in (\ref{h:leveln}), instead use
Bayes' rule to weigh interpretations based on a conjecture about speaker behavior. Speaker
behavior is regulated by a soft-max parameter $\lambda \geq 0$
\citep{luce:1959,sutton+barto:1998}. As $\lambda$ increases, choices approximate strict
maximization of expected utilities. Expected utility of a message $\messg$ in state $\state$
for a level $n+1$ speaker is here defined as $H_{n}(s|m;L)$, the probability that the hearer
will assign to or choose the correct meaning.\footnote{An alternative choice for the speaker's
  utility function, frequently made in Bayesian approaches after \citet{frank+goodman:2012}, is
  $\log H_{n}(s|m;L)$. Conceptually, the difference is that without the logarithm, utilities
  are derived from action choices, while with the logarithmic utilities are derived from how
  closely the listener's beliefs (after hearing an utterance) align with the speaker's, as
  measure by Kullback-Leibler divergence. In an evolutionary context, where fitness comes from
  actual interactions with the environment, not just solemn beliefs held about it, the
  formulation without the logarithm may be slightly more appealing. Moreover, log-based speaker
  utilities cause the speaker never to use false utterances. Since we aim here for a situation
  in which any type has a reasonable chance to be mistaken for any other based on their
  production behavior (see Section~\ref{sec:learnability}), we want speaker types which send
  any message in any state with positive probability.} For literal speakers, utility only
tracks truthfulness. Literal speakers choose any true message with equal probability but may
send false messages as well with a probability dependent on $\lambda$.

@@

In contrast to their literal counterparts of level $0$, pragmatic agents of level $1$ who use
$\Llack$ associate $\msome$ preferentially with $\ssome$. This association is not perfect, and
usually less strong than what agents with a lexicalized upper bound in $\Lbound$ can achieve --
with or without pragmatic reasoning. Higher order reasoning beyond level $1$ leads to stronger
associations of $\msome$ and $\ssome$ also for the receiver. Still, the case study presented in
Section~\ref{sec:si-case-study} will consider sender and receiver behavior at levels 0 and 1,
as the latter are the simplest pragmatic reasoning types which show a tendency to
communicatively attuned pragmatic enrichment.

@@

In biological evolution, where types are expressed genetically, transmission infidelity comes
into the picture through infrequent and mostly random mutation and genetic drift
\citep{kimura:1983}. However, an agent's lexicon and a disposition for pragmatic reasoning are
likely not inherited genetically. They need to be learned from observation, such as in
conditional imitation (see Appendix~\ref{sec:deriv-repl-mutat}).  Concretely, when agents learn
from or imitate type $j$, they observe the overt linguistic behavior of type $j$ and need to
infer the covert type that most likely produced the observed behavior.

@@

Following \citet{griffiths+kalish:2005,griffiths+kalish:2007} we model language acquisition as
a process of Bayesian inference in which learners combine the likelihood of a type producing
the witnessed learning input with prior inductive biases. In a Bayesian setting these biases
can be codified in a prior $P \in \Delta(T)$, which reflects the amount of data a learner
requires to faithfully acquire the language of the teacher
\citep[cf.][450]{griffiths+kalish:2007}.

The prior's influence depends on particulars of the learning or inference process. Early
simulation results suggested that weak biases could be magnified by exposing learners to only
small data samples \citep[e.g. in][]{brighton:2002}. \citet{griffiths+kalish:2007} proved that
if learners adopt a random type as a sample from their posterior beliefs about which types may
have generated the learning data that they saw, the population distribution of types will
eventually come to match the learners' prior distribution over types exactly. More
deterministic strategies such as the adoption of the type with the highest posterior
probability, so-called {\it maximum a posterior estimation} (MAP), increase the influence of
both the prior and the data \citep{griffiths+kalish:2007,kirby+etal:2007}. In the following, we
use a parameter $l\ge1$ to modulate between posterior sampling and the MAP strategy. When
$l = 1$ learners sample from the posterior. The learners' propensity to maximize the posterior
grows as $l$ increases.\footnote{The way the posterior over types is formed and how learners
  employ it are important design choices. There are other alternatives as well. For instance,
  instead of adopting a single type, learners could keep their uncertainty over types and base
  their production behavior on the posterior over types they form
  \citep{burkett+griffiths:2010}. A straightforward operationalization of this idea would then
  see production defined as a two-step process wherein a speaker first samples from her
  posterior and then produces utterances based on the type she sampled, as done in
  \citet{kirby+etal:2015}. Beyond adding complexity to the model, a disadvantage of this
  approach is that it could happen that a speaker that first uttered a Hungarian utterance
  speaks Quechua in the next, so to speak. We therefore maintain the simpler assumption of a
  single type being acquired in the present context, and postpone further investigation of more
  complex type-adoption schemes to future research.}

@@

When $l=1$ learners sample a type to adopt from their posterior distribution which they obtain
from observing the teacher produce utterances in particular states. From
\citet{griffiths+kalish:2007}, we know that under $l=1$ iterated Bayesian learning will
converge to a population proportion that exactly matches the agents' prior over types. If $l>1$
learners tend to pick types which are \emph{a posteriori} more likely with a higher probability
than less likely types. Even in the absence of a prior bias, iterated Bayesian learning with
$l>1$ can lead to populations in which a particular type $t_i$ has a higher proportion than
other types. For $l>1$, much hinges on particular asymmetries in production likelihoods
$P(d \mid t_j)$, and therefore on the composition of the set of types in general. Intuitively
put, if there are many types $t_j$ which all happen to (``erroneously'') produce data that is
most likely produced (``correctly'') by $t_i$, this can lead to a larger proportion of $t_i$ in
the long run even when priors over types are uniform. Appendix~\ref{sec:example-iter-bayes}
provides an abstract example for these likelihood driven effects.
